Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-20 doc/mc-paper.tex
Fine tuning

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi$, their first task is to
check whether $\phi \imply R$, i.e., that any model of $\phi$ is also
a model of $R$.  They then base other parts of their proof on that
property and use a separate checker to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it may
not preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_1 \lor \obar{x}_2)$\\    
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has five.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however the
second clause of $\phi_2$ is RAT with respect to the first---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2 = 0$.

This weakness would allow a buggy (or malicious) version
of \cdfour{} to generate a compiled representation of $\phi_2$ when
given $\phi_1$ as its input.  By adding the second clause to $R$,
the check with \dtrim{} would pass, as would the other tests performed
by their checker.  We have been able to confirm this possibility with
their compiler and checker\footnote{Downloaded May 18, 2023 from
\url{https://github.com/crillab/d4}}.

@@

In essence, a POG represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

@@

\section{The CPOG POG Representation and Proof System}
\label{sect:cpog}

@@

\section{Implementations}
We have implemented programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}\footnote{All tools, along with a set of 
benchmark problems are available at \url{https://github.com/rebryant/cpog}}.  The proof generator is the same in both
cases, since it need not be trusted.
Our \emph{verified}
version of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our \emph{prototype} version is written in C\@.
It is faster and
more scalable, but we anticipate its need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

>>>>

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi$, their first task is to
check whether $\phi \imply R$, i.e., that any assignment that satisfies $\phi$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate checker to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it may
not preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_1 \lor \obar{x}_2)$\\    
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has five.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however the
second clause of $\phi_2$ is RAT with respect to the first---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2 = 0$.

This weakness would allow a buggy (or malicious) version
of \cdfour{} to generate a compiled representation of $\phi_2$ when
given $\phi_1$ as its input.  By adding the second clause to $R$,
the check with \dtrim{} would pass, as would the other tests performed
by their checker.  We have been able to confirm this possibility with
their compiler and checker\footnote{Downloaded May 18, 2023 as\\
\url{https://github.com/crillab/d4/tree/333370cc1e843dd0749c1efe88516e72b5239174}.}

@@

A POG represents a partitioned-operation
formula with a sharing of common subformulas.  Every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

@@

\section{The CPOG Representation and Proof System}
\label{sect:cpog}

@@

\section{Implementations}
We have implemented programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}\footnote{The source code for all tools, along with a set of 
benchmark problems are available as part of the supplement\cite{bryant:sat:2023:supplement}}.  The proof generator is the same in both
cases, since it need not be trusted.
Our \emph{verified}
version of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our \emph{prototype} version is written in C\@.
It is faster and
more scalable, but we anticipate its need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.
